# Homeownership market {#sec-market-homeown}

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

# Load common script
source("_common.R")

# Load other packages other than tidyverse and hdatools
library(sf)
library(leaflet)
library(geojsonio)
library(classInt)
library(htmltools)
library(htmlwidgets)

```

### Existing Supply

There are a little over 101,000 homeownership units throughout the region.[^market-homeown-1] This data incorporates market trends from the last four years and the for-sale housing market has changed drastically in that time. Interest rates, which have the greatest potential to influence monthly mortgage payments, peaked at 7.79% in October 2023 and settled around 6.95% as of July 2024[^market-homeown-2]. While regional resales peaked in 2021, likely with much lower purchase prices and interest rates, more recent home sales with increased pricing and rates may influence these data by pulling monthly housing costs higher for all ownership units.

[^market-homeown-1]: This estimate includes units that are vacant, but “For sale only” and “Sold, not occupied”.

[^market-homeown-2]: FreddieMac Primary Mortgage Market Survey as of July 2024.

```{r}
#| label: fig-owner-supply
#| fig-cap: "Owner-occupied homes by monthly housing costs"

# Read in the selected monthly housing cost data for homeowners.
owner_supply <- read_rds("data/b25094_region.rds") |> 
  mutate(
    monthly = fct_relevel(
      monthly,
      c("Less than $599", "$600 to $999", "$1,000 to $1,499",
        "$1,500 to $1,999", "More than $2,000")
    )
  )

# sum(owner_supply$adj_estimate)

ggplot(
  owner_supply,
  aes(x = monthly, y = adj_estimate, label = label_comma()(adj_estimate))
  ) +
  geom_col(fill = hda_pal[1]) +
  geom_text(color = "white", nudge_y = -1500) +
  scale_y_continuous(labels = label_comma()) +
  theme_hda() +
  add_zero_line() +
  labs(
    title = "Owner-occupied homes by monthly housing costs",
    subtitle = "Inludes mortgage payments, property taxes, insurance, and basic utilities",
    caption = acs_cap("B25094")
  )

```

Most homes in the existing supply are single family units with three to four bedrooms, units that could easily accommodate households with upwards of four people and more. Around 15,000 units in the region have five bedrooms or more, and less than 10,000 homes in the region are sized for families of two people or fewer. This may indicate a market priority to larger and higher-income households who can afford both higher interest rates and increased bedroom sizes.

```{r}
#| label: fig-owner-bedroom
#| fig-cap: "Existing homeowner supply by bedrooms"

# The data visualization below shows the distribution of homeowner occupied housing
# units by the number of bedrooms in 2022. This viz is focused on the entire region. It 
# accounts for vacant housing units by applying an equal distribution of vacant 
# units based on the distribution of bedrooms.

# Read in the tenure by bedrooms data and filter for only homeowners.
owner_bedroom <- read_rds("data/b25042_region.rds") |> 
  filter(tenure == "Homeowner")

# Create a custom order to arrange the monthly cost values appropriately.
custom_order <- factor(owner_bedroom$br, levels = c("No bedroom", "1 bedroom", "2 bedrooms", 
                                                    "3 bedrooms", "4 bedrooms", "5 or more bedrooms"))
ggplot(owner_bedroom,
       aes(x = custom_order,
           y = adj_estimate)) +
  geom_col(position = "dodge", fill = "#445ca9") +
  theme_hda(base_size = 10) +
  scale_fill_hda() +
  scale_y_continuous(labels = number_format(big.mark = ",")) +
  labs(title = "Existing homeowner supply by bedrooms",
       caption = "Source: U.S. Census Bureau, 5-Year American Community Survey, Table B25042.")

```

. . .

### Building Trends

The region offers very little diversity of housing type, as single-family homes dominate both the rental and ownership markets. Of the existing ownership units region-wide, Stafford, Spotsylvania, and Fredericksburg respectively hold the largest shares of non-single family ownership units. Construction and current sales both overwhelmingly favor single family homes, and thus the existing supply reflects these trends in cost and size over time.

```{r}
#| label: fig-owner-type
#| fig-cap: "Existing homeowner supply by type and year built"

# The data visualization below shows the distribution of homeowner occupied housing
# units by structure type and year built in 2022. This viz is focused on the entire region. It 
# accounts for vacant housing units by applying an equal distribution of vacant 
# units based on the distribution of units by structure and year.

# Read in the tenure by structure and yearbuilt data and filter for only homeowners.
owner_type <- read_rds("data/b25127_region.rds") |> 
  filter(tenure == "Homeowner")

# Create a title text object with html/css for styling
title_text <- "<span style = 'color:#445ca9'><b>Single-family</b></span> housing has dominated construction"

# Create a title text object with html/css for styling
subtitle_text <- "<span style = 'color:#cce0ce'><b>Other</b></span> housing like manufactured housing follows, but still pales in comparison"

ggplot(owner_type,
       aes(x = yrbuilt,
           y = adj_estimate,
           fill = structure)) +
  geom_col() +
  theme_hda(base_size = 10) +
  scale_fill_hda(direction = -1) +
  scale_y_continuous(labels = number_format(big.mark = ",")) +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "Source: U.S. Census Bureau, 5-Year American Community Survey, Table B25127.")



```

*Note: This graph only shows homeownership units.*

Single-family construction has dominated over the past two decades, but the post-Great Recession recovery for the region has been slow. Permitting and construction data shows that the region has yet to recover to pre-Recession building levels with its height in 2005 for both rental and ownership unit development.

Construction appears to have been impacted by increasing interest rates and other inflationary pressures felt nation-wide in recent years, as permitting for single units has slowed following 2020. With a peak in single unit development in 2022 with about 2,500 permits that year, single family construction has decreased since then.

2024 may result in greater single family construction, to show recovery after interest rate increases in 2023, as the first quarter of 2024 shows about 1000 regional permits. If development continues steadily at this pace, construction for new single family units will nearly surpass pre-Great Recession levels.

. . .

```{r}
#| label: fig-bps-chart
#| fig-cap: "Building Permits"

# The data visualization below shows building trends broken out by type of housing in the FAAR region.

bps <- read_rds("data/faar_cbps.rds") |> 
  filter(NAMELSAD != "Orange County") |> 
  mutate(type = case_when(
    type == "2-units" ~ "2-4 units",
    type == "3-4 units" ~ "2-4 units",
    TRUE ~ type
  )) |> 
  group_by(year, type) |> 
  summarise(units = sum(units)) 



subtitle_text <- "However, <span style = 'color: #e9ab3f'><b>larger multifamily</b></span> construction has been on the rise."

ggplot(bps, 
       aes(x = year,
           y = units,
           fill = type)) +
  geom_col() +
  facet_wrap(~type, nrow = 1) +
  scale_fill_hda() +
  theme_hda(base_size = 10) +
  labs(title = "The region has still not recovered to pre-Great Recession construction levels",
       subtitle = subtitle_text,
       caption = "Source: Census Building Permits Survey")
  

```

*Note: This table includes permitting for both rental and ownership units.*

### Market Trends

The regional for-sale market is relatively strong, with core hot spots present where resales happen most quickly. Most single family home sales in the past four years have been along the 95 Corridor in Fredericksburg City and in Spotsylvania and Stafford counties. The map below shows the number of residential home sales between January 2020 through April 2024, with dark hexagons representing a higher number of home sales.

```{r}

#| label: fig-price-line
#| fig-cap: "Median sales price"

mls_msp <- read_rds("data/faar_mls.rds") 
  #filter(county != "Orange") |> 
  select(quarter, type, adj_sales) |> 
  st_drop_geometry() |> 
  group_by(quarter, type) |> 
  summarise(med_sales = median(adj_sales))


title_text <- "<span style = 'color:#445ca9'><b>New Construction</b></span> and <span style = 'color:#8baeaa'><b>Resale</span>"

ggplot(sales_price,
       aes(x = quarter,
           y = med_sales,
           color = type)) +
  geom_line(size = 1) +
  geom_point() +
  scale_color_hda() + 
  theme_hda(base_size = 10) +
  labs(title = title_text,
       subtitle = "In 2024 Q2 inflation-adjusted dollars",
       caption = "Source: BrightMLS; CPI for All Urban Consumers: All Items Less Shelter in U.S. City Average") +
  scale_y_continuous(labels = dollar_format())

```

Additionally, as these data incorporate market trends from the last four years, and the for-sale housing market has changed drastically in that time, especially in more recent years. Interest rates, which have the greatest potential to influence monthly mortgage payments, have peaked at 7.79% in October 2023 and settled around 6.95% as of July 2024.[^market-homeown-3] While regional resales peaked in 2021, likely with much lower purchase prices and interest rates, recent home sales with increased pricing and rates may influence these data by pulling monthly housing costs for all ownership units higher.

[^market-homeown-3]: FreddieMac Primary Mortgage Market Survey as of July 2024.

```{r}
#| label: fig-owner-costs
#| fig-cap: "Monthly Owner Costs and Interest Rates"


```

```{r}
#| label: fig-sales-map-nc
#| fig-cap: "Map of total new construction home sales from 2020 through April 2024"

# Create a hex grid map showing home sale concentrations in FAAR region.



# sales <- read_rds("data/faar_mls_cpi_google.rds") |>
#   mutate(longitude = case_when(
#     mls_number == "630778" ~ -78.329369,
#     TRUE ~ longitude)) |>
#   mutate(latitude = case_when(
#     mls_number == "630778" ~ 38.214775,
#     TRUE ~ latitude)) |>
#   mutate(longitude = case_when(
#     mls_number == "VAOR2002600" ~ -78.181042,
#     TRUE ~ longitude)) |>
#   mutate(latitude = case_when(
#     mls_number == "VAOR2002600" ~ 38.137717,
#     TRUE ~ latitude)) |>
#   filter(mls_number != "606389") |>  # This entry is incorrectly marked as being in Spotsylvania County
#   st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
# 
# 
# write_rds(sales, "data/faar_mls_check.rds")
# geojson_write(sales, file = "data/mls.geojson")
# 


sales <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |> 
  filter(structure_type != "Garage/Parking Space") |> 
  filter(new_construction_yn == "Yes")

# leaflet(sales) |> 
#   addProviderTiles(providers$CartoDB.Positron) |> 
#   addCircleMarkers()

hex_grid <- st_make_grid(sales, n = c(50,50),
                         what = "polygons", square = FALSE)

hex_grid_sf <- st_sf(hex_grid) %>% 
  mutate(grid_id = 1:length(lengths(hex_grid)))

hex_grid_sf$n_pts <- lengths(st_intersects(hex_grid_sf, sales))

hex_count <- filter(hex_grid_sf, n_pts > 0)

# Define the base color
base_color <- "#445ca9"

# Create a color palette function based on the base color
color_palette <- colorRampPalette(colors = c("#cce0ce", base_color))

# Number of bins (adjust as needed)
bins <- 9

# Generate the color palette for the bins
palette_colors <- color_palette(bins)

# color_bin_palette <- colorBin(palette = palette_colors, domain = hex_count$n_pts, bins = bins)


# palette <- colorBin('Greens', domain = hex_count$n_pts, bins = 9)

quantiles <- classIntervals(hex_count$n_pts, n=5, style='jenks')
palette_quantiles <- colorBin(palette_colors, domain = hex_count$n_pts, bins=quantiles$brks, pretty=FALSE)


leaflet(hex_count) |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~palette_quantiles(n_pts),
    fillOpacity = 0.7,
    opacity = 1,
    weight = 1,
    color='white',
    label = ~as.character(n_pts),  # Show n_pts value when hovered over
    highlightOptions = highlightOptions(
      weight = 3,
      color = '#666',
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    popup = ~paste("n_pts:", n_pts)  # Show n_pts value on click
  ) |> 
  addLegend(
    pal = palette_quantiles,
    values = ~n_pts,
    title = "Number of Sales",
    position = "bottomleft"
  )


```

```{r}
#| label: fig-sales-map-resale
#| fig-cap: "Map of total homes resales from 2020 through April 2024"

# Create a hex grid map showing home sale concentrations in FAAR region.

library(leaflet)
library(sf)
library(geojsonio)
library(classInt)
library(htmltools)
library(htmlwidgets)

# sales <- read_rds("data/faar_mls_cpi_google.rds") |>
#   mutate(longitude = case_when(
#     mls_number == "630778" ~ -78.329369,
#     TRUE ~ longitude)) |>
#   mutate(latitude = case_when(
#     mls_number == "630778" ~ 38.214775,
#     TRUE ~ latitude)) |>
#   mutate(longitude = case_when(
#     mls_number == "VAOR2002600" ~ -78.181042,
#     TRUE ~ longitude)) |>
#   mutate(latitude = case_when(
#     mls_number == "VAOR2002600" ~ 38.137717,
#     TRUE ~ latitude)) |>
#   filter(mls_number != "606389") |>  # This entry is incorrectly marked as being in Spotsylvania County
#   st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
# 
# 
# write_rds(sales, "data/faar_mls_check.rds")
# geojson_write(sales, file = "data/mls.geojson")
# 


sales <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |> 
  filter(structure_type != "Garage/Parking Space") |> 
  filter(new_construction_yn != "Yes")

# leaflet(sales) |> 
#   addProviderTiles(providers$CartoDB.Positron) |> 
#   addCircleMarkers()

hex_grid <- st_make_grid(sales, n = c(50,50),
                         what = "polygons", square = FALSE)

hex_grid_sf <- st_sf(hex_grid) %>% 
  mutate(grid_id = 1:length(lengths(hex_grid)))

hex_grid_sf$n_pts <- lengths(st_intersects(hex_grid_sf, sales))

hex_count <- filter(hex_grid_sf, n_pts > 0)

# Define the base color
base_color <- "#445ca9"

# Create a color palette function based on the base color
color_palette <- colorRampPalette(colors = c("#cce0ce", base_color))

# Number of bins (adjust as needed)
bins <- 9

# Generate the color palette for the bins
palette_colors <- color_palette(bins)

# color_bin_palette <- colorBin(palette = palette_colors, domain = hex_count$n_pts, bins = bins)


# palette <- colorBin('Greens', domain = hex_count$n_pts, bins = 9)

quantiles <- classIntervals(hex_count$n_pts, n=5, style='jenks')
palette_quantiles <- colorBin(palette_colors, domain = hex_count$n_pts, bins=quantiles$brks, pretty=FALSE)


m <- leaflet(hex_count) |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~palette_quantiles(n_pts),
    fillOpacity = 0.7,
    opacity = 1,
    weight = 1,
    color='white',
    label = ~as.character(n_pts),  # Show n_pts value when hovered over
    highlightOptions = highlightOptions(
      weight = 3,
      color = '#666',
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    popup = ~paste("n_pts:", n_pts)  # Show n_pts value on click
  ) |> 
  addLegend(
    pal = palette_quantiles,
    values = ~n_pts,
    title = "Number of Sales",
    position = "bottomleft"
  )


```

```{r}
#| label: type-bar
#| fig-cap: "Total sales"

# Create a bar chart showing the number of home sales per quarter. Fill bar stacks based on jurisdiction.

sales_type <- read_rds("data/faar_mls_check.rds")|> 
  filter(county != "Orange, VA") |> 
  mutate(type = case_when(
    new_construction_yn == "Yes" ~ "New Construction",
    TRUE ~ "Resale"
  )) |> 
  filter(structure_type != "Garage/Parking Space") |> 
  mutate(structure_type = case_when(
    structure_type == "End of Row/Townhouse" ~ "Townhouse",
    structure_type == "Interior Row/Townhouse" ~ "Townhouse",
    structure_type == "Penthouse Unit/Flat/Apartment" ~ "Unit/Flat/Apartment",
    structure_type == "Mobile Pre 1976" ~ "Other",
    TRUE ~ structure_type
  )) |> 
  group_by(structure_type, type, quarter) |> 
  summarise(count = n_distinct(mls_number))

# title_text <- "<span style = 'color:#445ca9'><b>New Construction</b></span> and <span style = 'color:#8baeaa'><b>Resale</span> by quarter"

ggplot(sales_type,
       aes(x = quarter,
           y = count,
           fill = reorder(structure_type, -count))) +
  geom_col(position = "stack") +
  facet_wrap(~type) +
  scale_fill_hda() +
  theme_hda(base_size = 10) +
  labs(title = "Residential sales by sale type and structure type",
       caption = "Source: BrightMLS") +
  theme(legend.position = "right") +
  scale_y_continuous(labels = number_format(big.mark = ","))

```

```{r}
#| label: dom-bar
#| fig-cap: "Days on Market"

# Create a bar chart showing the number of home sales per quarter. Fill bar stacks based on jurisdiction.

sales_dom <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |>
  mutate(type = case_when(
    new_construction_yn == "Yes" ~ "New Construction",
    TRUE ~ "Resale"
  )) |> 
  filter(structure_type != "Garage/Parking Space") |> 
  select(quarter, cdom, type) |> 
  filter(cdom != 0) |> 
  group_by(quarter, type) |> 
  summarise(med_dom = median(cdom))

title_text <- "<span style = 'color:#445ca9'><b>New Construction</b></span> and <span style = 'color:#8baeaa'><b>Resale</span> median days on market"

ggplot(sales_dom,
       aes(x = quarter,
           y = med_dom,
           fill = type)) +
  geom_col(position = "dodge") +
  theme_hda(base_size = 10) +
  scale_fill_hda() +
  labs(title = title_text,
       caption = "Source: BrightMLS")
```

```{r}
#| label: bed-plot
#| fig-cap: "Bedrooms"

# Create a data viz that shows the distribution of home sales based on bedroom count.


beds <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |> # Filter out Orange County
  mutate(type = case_when( # Create a label for New Construction and Resale
    new_construction_yn == "Yes" ~ "New Construction",
    TRUE ~ "Resale"
  )) |> 
  filter(structure_type != "Garage/Parking Space", # Remove garage and parking space sales
         beds != 0) |>  # Remove home sales wherein there are no bedrooms listed due to incomplete entries
  mutate(beds = case_when( # Group bedrooms of 5 or more
    beds > 5 ~ 5, 
    TRUE ~ beds
  )) |> 
  mutate(beds = as.character(beds)) |> # Transform beds into character rather than a numeric
  mutate(beds = case_when( # Transform the 5 count to "5 or more"
    beds == "5" ~ "5 or more",
    TRUE ~ beds
  )) |> 
  mutate(count = 1) |> # Create a column to help calculate a sum
  group_by(type, beds) |> 
  summarise(sum = sum(count)) |> 
  ungroup() |>
  group_by(type) |> 
  mutate(pct = sum/sum(sum))


ggplot(beds, 
       aes(x = beds,
           y = pct,
           fill = type)) +
  geom_col(position = "dodge") +
  theme_hda(base_size = 10) +
  labs(title = "Distribution of Bedrooms",
       subtitle = "Percent of total by sales type",
       caption = "Source: BrightMLS, excludes listings where bedrooms are listed as zero.") +
  facet_wrap(~type, nrow = 1) +
  scale_fill_hda() +
  scale_y_continuous(labels = percent_format())



```

```{r}
#| label: sf-plot
#| fig-cap: "Square Footage"

# Create a data viz that shows the distribution of home sales based on square footage.
# Density plot more easily handles outliers.

sales_sf <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |> 
  mutate(type = case_when(
    new_construction_yn == "Yes" ~ "New Construction",
    TRUE ~ "Resale"
  )) |> 
  filter(structure_type != "Garage/Parking Space") |> 
  select(total_finished_sqft, type) |> 
  filter(total_finished_sqft != 0) |> 
  filter(total_finished_sqft < 5000)

subtitle_text <- "<span style = 'color:#445ca9'><b>New Construction</b></span> and <span style = 'color:#8baeaa'><b>Resale</span>"


ggplot(sales_sf, 
       aes(x = total_finished_sqft,
           fill = type)) +
  geom_density(
    alpha = 0.6,
    color = "white"
  ) +
  theme_hda(base_size = 10) +
  labs(title = "Distribution of Square Footage",
       subtitle = subtitle_text,
       x = "Square Footage",
       y = "Density",
       caption = "Source: BrightMLS, excludes listings where sq. ft. is listed as zero or above 5,000.") +
  scale_fill_hda() +
  scale_y_continuous(labels = number_format())


```

```{r}
#| label: lot-plot
#| fig-cap: "Lot Size"

# Create a data viz that shows the distribution of home sales based on lot size.
# Density plot more easily handles outliers.

sales_lot <- read_rds("data/faar_mls_check.rds") |> 
  filter(county != "Orange, VA") |> 
  mutate(type = case_when(
    new_construction_yn == "Yes" ~ "New Construction",
    TRUE ~ "Resale"
  )) |> 
  filter(structure_type != "Garage/Parking Space") |> 
  select(lot_size_acres, type) |> 
  filter(lot_size_acres != 0) |> 
  filter(lot_size_acres < 10)

title_text <- "<span style = 'color:#011E41'><b>New Construction</b></span> and <span style = 'color:#40C0C0'><b>Resale</span> by quarter"

ggplot(sales_lot, 
       aes(x = lot_size_acres,
           fill = type)) +
  geom_density(
    alpha = 0.6,
    color = "white"
  ) +
  theme_hda(base_size = 10) +
  labs(title = "Distribution of Lot Size",
       subtitle = subtitle_text,
       x = "Acres",
       y = "Density",
       caption = "Source: BrightMLS, excludes listings where lot size is listed as zero or above 10 acres.") +
  scale_fill_hda() +
  scale_y_continuous(labels = number_format())

```
