---
title: "needs-current"
format: html
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

source("r/affordcalc.R")
source("_common.R")
library(lubridate)


#| label: income-costs-data
#| eval: false

# Load income data

Incomes <- read_rds("data/qcew_data_cpi.rds")

average_annual <- Incomes |> 
  rename(dollars = estimate,
         locality = NAME) |> 
  mutate(locality = str_remove(locality, ", Virginia")) |>
  filter(tenure == "Renter") |> 
  select(locality, year, tenure, dollars) 


# Load rent data

faar_rent_annual <- read_rds("data/faar_rent_ann.rds") |>
  filter(locality %in% c("Stafford County", 
                        "Spotsylvania County", 
                        "City of Fredericksburg", 
                        "King George County", 
                        "Caroline County",
                        "Orange County"))


# Load sales data and summarize median prices by locality and year

faar_mls_annual <- read_rds("data/faar_mls.rds") |> 
  mutate(year = as.numeric(year(close_date))) |> 
  rename(locality = county) |> 
  summarise(
    msp = median(close_price),
    .by = c(locality, year)
  ) |> 
  mutate( locality = case_when(
  locality == "Caroline" ~ "Caroline County",
  locality == "King George" ~ "King George County",
  locality == "Orange" ~ "Orange County",
  locality == "Spotsylvania" ~ "Spotsylvania County",
  locality == "Stafford" ~ "Stafford County",
  locality == "Fredericksburg" ~ "City of Fredericksburg"
  )
  )


# Join rents and sales to incomes

costs_income <- faar_rent_annual |> 
  mutate(
    locality = str_remove(locality, ", Virginia"),
    rent = avg_price  # simplified since avg_price is already numeric
  ) |>
  left_join(renter_income, by = c("locality", "year")) |> 
  filter(year > 2020) |>
  select(locality, year, dollars, rent) |>
  rename(income = dollars) |>
  left_join(faar_mls_annual, by = c("locality", "year"))


# Add affordable rents and sales prices

int_rates <- fredmac_int_annual

downpayment <- 0.05 # 5% downpayment
closingcosts <- 0.015 # 1.5% closing costs
taxes <- 250 # Assume $250/month for taxes
ins_fees <- 150 # Assume $150/month for insurance and fees

price_calc <-  function(x, y) {
  
  monthly_cost <- x / 12 * 0.28
  monthly_pmt <- monthly_cost - (taxes + ins_fees)
  principal <- monthly_pmt * (((1-1/(1+y)^360))/(y/12))
  loan_amt <- principal/(1-closingcosts)
  price <- (loan_amt * -1) / ((downpayment) - 1)

  price
  
}

costs_income_afford <- costs_income |> 
  left_join(int_rates, by = "year") |> 
  mutate(rent_afford = income / 12 * 0.3) |> 
  mutate(price_afford = price_calc(income, annual_int))

write_rds(costs_income_afford, "data/costs_income_afford.rds")



#| label: income-costs-change
#| eval: false

costs_income_change <- costs_income |> 
  group_by(locality) |> 
  arrange(year, .by_group = TRUE) |> 
  mutate(
    rent_chg = (rent - lag(rent)),
    price_chg = (avg_price - lag(avg_price)),
    inc_chg = (income - lag(income))
  ) |> 
  mutate(
    across(.cols = c(rent_chg, price_chg, inc_chg), ~replace_na(.x, 0))
  ) |> 
  mutate(
    across(.cols = c(rent_chg, price_chg, inc_chg), ~cumsum(.x)),
    rent_pct = rent_chg/first(rent),
    price_pct = price_chg/first(avg_price),
    inc_pct = inc_chg/first(income)
  ) |> 
  select(locality, year, rent_pct, price_pct, inc_pct) |>
  pivot_longer(
    cols = c(rent_pct, price_pct, inc_pct),
    names_to = "type",
    values_to = "pct_chg"
  ) |> 
  mutate(
    pct_chg = case_when(
      year > 2020 & type == "inc_pct" ~ NA_real_,
      TRUE ~ pct_chg
    ),
    type = case_when(
      type == "inc_pct" ~ "Income",
      type == "price_pct" ~ "Home price",
      type == "rent_pct" ~ "Rent",
      TRUE ~ type
    )
  )


write_rds(costs_income_change, "data/costs_income_change.rds")


```

# Housing Needs Analysis

This analysis's goal is to show existing housing costs versus the ability of households in the region to afford those housing costs. Along with comparing supply-side factors in the housing market study, this analysis will pinpoint any obstacles or worker shortages in the area and focus on connect supply and demand data. In order to support the working population, this research will conclude with broad estimates and analysis of the region's anticipated needs for the ensuing ten years.

## Overview

\[adding in JTK's new text and spectrum pieces\]

At a glance, we can see supply and demand in the region are not in alignment. There is an excess of supply for \[ XYZ and a lack of supply for XYZ. \]

```{r}
remotes::install_github("hdadvisors/hdatools")
library(hdatools)
install.packages("gtable")

supply_demand <- read_rds("data/fxburg_supply_demand.rds")

colnames(supply_demand)
glimpse(supply_demand)

## Separate Renter and Homeowner S/D

supply_demand_renter <- supply_demand |>
  filter(tenure == "Renter") |>
  group_by(ami, value) |>
  summarize(
    estimate = sum(estimate),
    .groups = "drop"
  )
  
supply_demand_homeowner <- supply_demand |>
  filter(tenure == "Homeowner") |>
  group_by(ami, value) |>
  summarize(
    estimate = sum(estimate),
    .groups = "drop"
  )
  
# Set factors for AMIs

ami_levels <- c("30% AMI or below", "31-50% AMI",
              "51-80% AMI", "81-100% AMI", "101-120% AMI",
              "121% AMI and greater")


# Add the factor directly to your filtered data
supply_demand <- supply_demand |>
  mutate(ami_order = factor(ami, levels = ami_levels))

supply_demand_renter <- supply_demand_renter |>
  mutate(ami_order = factor(ami, levels = ami_levels))

supply_demand_homeowner <- supply_demand_homeowner |> 
  mutate(ami_order = factor(ami, levels = ami_levels))


# Plot 

title_text <- "<span style = 'color:#8baeaa'><b>Supply</b></span> and <span style = 'color:#445ca9'><b>Demand</span> in the region for the Past X Years"


ggplot(supply_demand,
       aes(x = ami_order,
           y = estimate,
           fill = value)) +
  geom_col(position = "dodge") +
  facet_wrap(~tenure) +
  theme_hda(base_size = 10) +
  scale_fill_hda() +
  coord_flip() +
  labs(title = title_text)


```

## Incomes vs Housing Costs

\[ insert analysis \]

```{r}

costs_income_afford <- read_rds("data/costs_income_afford.rds")
costs_income_change <- read_rds("data/costs_income_change.rds")

library(ggiraph)
library(scales)
library(ggplot2)

#| label: fig-costs-inc-change
#| fig.cap: "Cumulative percent change in median renter income, home prices, and average rent"
#| 

cost_plot <- ggplot(costs_income_change,
       aes(x = year,
           y = pct_chg,
           color = type)) +
  geom_line(size = 1) +
  geom_point_interactive(
    aes(data_id = pct_chg,
        tooltip = percent_format(accuracy = 0.1)(pct_chg)),
    size = 0.5) + 
  theme_pha() +
  scale_color_pha() + 
  facet_wrap(~locality, nrow = 1) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Cumulative percent change in median renter income,<br>home prices, and average rent",
       subtitle ="2020 to 2024 (ACS income data through 2020)",
       color = "Percent change",
       caption = "**Sources:** U.S. Census Bureau, American Community Survey, Table B25119;<br>Central Virginia MLS; CoStar Group, Inc.") +
  theme(legend.position = "bottom")

if (knitr::is_html_output()) {
  
  girafe(ggobj = cost_plot,
         height_svg = 4) 
  
} else {
  
  cost_plot
  
}


```

### Rental Housing Gap

```{r}
## Supply and Demand Renters by AMI 

title_text <- "<span style = 'color:#8baeaa'><b>Supply</b></span> and <span style = 'color:#445ca9'><b>Demand</span> in the region for Renters"


ggplot(supply_demand_renter,
       aes(x = ami_order,
           y = estimate,
           fill = value)) +
  geom_col(position = "dodge") +
  theme_hda(base_size = 10) +
  scale_fill_hda() +
  coord_flip() +
  labs(title = title_text)
```

```{r}
## NEED TO UPDATE TO ADD MORE YEARS

#| label: gap-data
#| eval: false

library(janitor)
faar <- c("51630", "51033", "51099", "51177", "51179", "51137") # FAAR FIPS codes
  
  
tb_18c <- read_csv("data/Table18C_2019to2020.csv")

colnames(tb_18c)[16] <- "Cost"

renter <- tb_18c |>
  clean_names() |>
  filter(line_type == "Detail") |>
  select(county, fips, year, estimate, tenure, cost, household_income) |>
  group_by(county, fips, year, tenure, cost, household_income) |>
  summarise(estimate = sum(estimate)) |>
  filter(fips %in% faar)

tb_18_match <- renter |>
  filter(tenure == "Renter occupied") |>
  mutate(cost = case_when(
    cost == "greater than RHUD30 and less than or equal to RHUD50" ~ "31 to 50 percent AMI",
    cost == "greater than RHUD50 and less than or equal to RHUD80" ~ "51 to 80 percent AMI",
    cost == "greater than RHUD80" ~ "80 percent AMI or greater",
    cost == "less than or equal to RHUD30" ~ "30 percent AMI or below"
  )) |>
  mutate(household_income = case_when(
    household_income == "greater than 100% of HAMFI" ~ "80 percent AMI or greater",
    household_income == "greater than 80% of HAMFI but less than or equal to 100% of HAMFI" ~ "80 percent AMI or greater",
    household_income == "greater than 50% of HAMFI but less than or equal to 80% of HAMFI" ~ "51 to 80 percent AMI",
    household_income == "greater than 30% of HAMFI but less than or equal to 50% of HAMFI" ~ "31 to 50 percent AMI",
    household_income == "less than or equal to 30% of HAMFI" ~ "30 percent AMI or below"
  )) |>
  group_by(county, fips, year, cost, household_income) |>
  summarise(estimate = sum(estimate))

tb_18_match <- tb_18_match |>
  mutate(match = case_when(
    cost == household_income ~ "Affordable",
    cost == "30 percent AMI or below" & household_income == "31 to 50 percent AMI" ~ "Very affordable",
    cost == "30 percent AMI or below" & household_income == "51 to 80 percent AMI" ~ "Very affordable",
    cost == "30 percent AMI or below" & household_income == "80 percent AMI or greater" ~ "Very affordable",
        cost == "31 to 50 percent AMI" & household_income == "31 to 50 percent AMI" ~ "Affordable",
    cost == "31 to 50 percent AMI" & household_income == "51 to 80 percent AMI" ~ "Very affordable",
    cost == "31 to 50 percent AMI" & household_income == "80 percent AMI or greater" ~ "Very affordable",
        cost == "31 to 50 percent AMI" & household_income == "30 percent AMI or below" ~ "Unaffordable",
    cost == "51 to 80 percent AMI" & household_income == "30 percent AMI or below" ~ "Unaffordable",
        cost == "51 to 80 percent AMI" & household_income == "31 to 50 percent AMI" ~ "Unaffordable",
    cost == "51 to 80 percent AMI" & household_income == "80 percent AMI or greater" ~ "Very affordable",
    cost == "80 percent AMI or greater" & household_income == "30 percent AMI or below" ~ "Unaffordable",
    cost == "80 percent AMI or greater" & household_income == "31 to 50 percent AMI" ~ "Unaffordable",
    cost == "80 percent AMI or greater" & household_income == "51 to 80 percent AMI" ~ "Unaffordable"
  )) |>
  mutate(gapcode = case_when(
    match == "Unaffordable" ~ "Gap",
    TRUE ~ "Matches or less than income"
  ))

write_rds(tb_18_match, "data/tb_18_match.rds")

tb_18_match <- read_rds("data/tb_18_match.rds")

gap <- tb_18_match |>
  filter(fips %in% as.numeric(faar)) |> 
  group_by(year, household_income, gapcode) |>
  summarise(estimate = sum(estimate)) |>
  mutate(estimate = case_when(
    gapcode == "Gap" ~ estimate * -1,
    gapcode == "Matches or less than income" ~ estimate
  )) |>
  filter(household_income != "80 percent AMI or greater")

write_rds(gap, "data/gap.rds")

```

[Comprehensive Housing Affordability Strategy (CHAS)](https://www.huduser.gov/portal/datasets/cp.html) data provided by the Department of Housing and Urban Development (HUD) allows us to understand the cost of housing in relation to household incomes. For renters making less than 80% AMI across the region, there has been little change in the gap in affordable rental housing.

```{r}

## UPDATE AS ABOVE

#| label: fig-gap-plot
#| fig.cap: "Rental housing gap by AMI"

gap |>
  mutate(gapcode = case_when(
    gapcode == "Gap" ~ "Unaffordable",
    TRUE ~ "Affordable"
  ))

gap_plot <- ggplot(gap,
       aes(x = year, 
           y = estimate,
           fill = str_wrap(gapcode, 25),
           data_id = estimate,
           tooltip = label_comma()(estimate))) +
  geom_col() + 
  facet_grid(~household_income) + 
  theme_hda() +
  scale_fill_hda() + 
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Rental housing gap by AMI",
       subtitle = "2015 to 2019",
       caption = "**Source:** U.S Department of Housing and Urban Development,<br>Comprehensive Housing Affordability Strategy (CHAS), Table 18C.") +
  theme(legend.position = "bottom") +
  add_zero_line("y")

if (knitr::is_html_output()) {
  
  girafe(ggobj = gap_plot,
         height_svg = 4) 
  
} else {
  
  gap_plot
  
  }

```

The current rental supply shows a large gap of units available for under \$1,000 a month, with only 5,000 units at that price point. The vast majority of rental units (over 18,000) cost more than \$1,500 per month, representing 57% of the total rental supply.

The limited availability of rental units for households who can afford \$2000 monthly could indicate that many higher income households are instead buying in that price point, where there are more units available, further driving demand for ownership units priced at \$2000 monthly (reflected in the number of ownership units at that price point).

More likely though, middle to higher income renting-households are able to take advantage of the greater availability of units priced between \$1000 and \$1999 (18,000 units compared with 7,500 for \$2,000 monthly). This creates increased competition for the “middle-priced” units on the regional rental market, driving cost-burden on low income households and opportunity for savings for higher income households.

## Rental Affordability

::: callout-tip
#### Calculating rental affordability

In this chapter, an "affordable rent" is no more than 30 percent of a household's gross monthly income. Any rent amount higher than this level would make the renter cost-burdened.
:::

Occupation data shows the most common salaries in the region, which influences what individuals can purchase or rent, where they must live depending on their wage, and how their families can effectively expand. In a [previous chapter](spectrum-work.qmd###Common%20industries%20and%20occupations) , the five most common occupation categories in the region were determined from the latest PUMS data.

The mismatch in salaries with average rental costs particularly affects workers in sectors such as food preparation, office administration, and some public sector jobs (e.g., teachers in Caroline County, social workers in Stafford and Fredericksburg) who can only afford rents around \$825-\$1,250 per month. Based on the number of homes for sale and rentals with higher monthly costs addressed in the market analysis, the supply of housing is not meeting the needs of over 500,000 regional job holders.

It is reasonable to assume then that many workers in Office and Administrative Support and Food Preparation and Serving Related Occupations are commuting into the region to work and living elsewhere. With affordable rent at \$1,159 and \$825.88 respectively, if these workers do live in the region, they are likely to be cost burdened.

```{r}

#| label: fig-rent-afford
#| fig.cap: "Average rents versus affordable rents for median renter income"

rent_afford <- costs_income_afford |> 
  select(1, 2, 4, 7) |> 
  pivot_longer(cols = 3:4,
               names_to = "type") |> 
  mutate(type = case_when(
    type == "rent" ~ "Average rent",
    TRUE ~ "Affordable rent"
  ))

rent_gap <- ggplot(rent_afford,
       aes(x = year,
           y = value,
           fill = type,
           data_id = value,
           tooltip = dollar_format()(value))) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_pha() +
  scale_fill_pha() +
  facet_wrap(~locality, nrow = 1) +
  scale_x_continuous(n.breaks = 7) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Average rents versus affordable rents<br>for median renter income",
       subtitle ="2016 to 2022 (ACS income data through 2020)",
       caption = "**Sources:** U.S. Census Bureau, American Community Survey, Table B25119; CoStar Group, Inc.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5))


if (knitr::is_html_output()) {
  
  girafe(ggobj = rent_gap,
         height_svg = 4) 
  
} else {
  
  rent_gap
  
  }

```

Fire professions in Stafford, Spotsylvania, and Fredericksburg have the most generous average incomes of the public professions analyzed for this study. Proportionally, this means these workers can afford more generous monthly housing costs, ranging from \$2000 to \$2500 in rent. On the other end of the spectrum, social workers in Stafford and Fredericksburg, as well as teaching professions in Caroline county, can only afford around \$1250 per month.

```{r}

locality_wages <- read_csv("data/faar_locality_wages.csv") |> 
  filter(
    job_actual != "Fire Protection",
    !(locality == "Orange" & job_actual == "Social Assistance")
  )


rate <- 0.065  # 6.5% annual interest rate
tax_ins <- 0.25  # Property tax and insurance as percentage of monthly payment (25%)
dp <- 0.20  # Down payment percentage (20%)


local_wages_cty <- locality_wages |> 
  summarize(
    salary = mean(salary), .by = locality
  ) |> 
  mutate(
    aff_rent = salary/12*0.3,
    aff_own = 
      (salary/12*0.28) /
      ((rate/12 * (1 + rate/12)^360) /
         ((1+rate/12)^360 - 1) /
         (1-(tax_ins/12)) /
         (1-dp))
  )

local_wages_cty |> 
  ggplot(aes(x = aff_rent, y = fct_reorder(locality, aff_rent), label = label_dollar(accuracy = 1)(aff_rent))) +
  geom_col(fill = hda_pal[1]) +
  geom_text(color = "white", hjust = 0.5, nudge_x = -100) +
  geom_vline(xintercept = 1620) +
  geom_vline(xintercept = 1812) +
  geom_text(
    aes(x = 1620, y = "King George", label = "Average\n1-bed\nrent:\n$1,620"), vjust = 0.6, hjust = 0, nudge_x = 20
  ) +
  geom_text(
    aes(x = 1812, y = "King George", label = "Average\n2-bed\nrent:\n$1,812"), vjust = 0.6, hjust = 0, nudge_x = 20
  ) +
  scale_x_continuous(limits = c(0, 1950), expand = c(0.01, 0.01), labels = label_dollar()) +
  scale_y_discrete(expand = c(0.1, 0)) +
  add_zero_line("x") +
  labs(
    title = "Affordable rents for public sector salaries",
    subtitle = "Based on average wages for police, fire, and teacher jobs",
    caption = "**Sources:** HDAdvisors calculations of local government salary data; CoStar Group, Inc."
  ) +
  theme_hda(flip_gridlines = T)


```

::: callout-tip
Note the Bureau of Labor Statistics does not differentiate where workers live, or their locality of origin. These data outline only where the profession is based.
:::

From the [previous-chapter](market-rental.qmd#rentalmarket), we know that there are xxx rental units in the region priced at \$1,500 and below.

\[more analysis\]

## Attaining Homeownership

::: callout-tip
To determine how affordable homeownership is at the median renter income, we can calculate the maximum home sales price affordable to a buyer with that income. To make these estimates, we make the following simplified assumptions:

-   5 percent down payment
-   1.5 percent in closing costs
-   \$250 per month for property taxes
-   \$150 per month for insurance and other costs

For underwriting purposes, we assume that the monthly mortgage payment plus these costs can not exceed 28 percent of the buyer’s gross income. For example, a renter earning \$50,000 can afford a monthly housing cost no more than \$1,166.67. To determine the maximum principal amount, and the subsequent sales price, we assume a standard 30-year fixed-rate mortgage using the average annual interest rates published by Freddie Mac
:::

```{r}
#| label: fig-price-afford
#| fig.cap: "Median sales price versus maximum home price affordable to median renter income"

price_afford <- costs_income_afford |> 
  select(1, 2, 5, 8) |> 
  pivot_longer(cols = 3:4,
               names_to = "type") |> 
  mutate(type = case_when(
    type == "price" ~ "Median sales price",
    TRUE ~ "Maximum affordable home price"
  ))

slgap <- ggplot(price_afford,
       aes(x = year,
           y = value,
           fill = type,
           data_id = value,
           tooltip = dollar_format()(value))) +
  geom_col(position = "dodge") +
  geom_col_interactive(position = "dodge", size = 2) +
  theme_hda() +
  scale_fill_hda() +
  facet_wrap(~locality, nrow = 1) +
  scale_x_continuous(n.breaks = 7) +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Median sales price versus <br>maximum home price affordable to median renter income",
       subtitle ="2016 to 2022 (ACS income data through 2020)",
       caption = "**Sources:** U.S. Census Bureau, American Community Survey,<br>Table B25119 and CVR MLS.") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5))

if (knitr::is_html_output()) {
  
  girafe(ggobj = slgap,
         height_svg = 4) 
  
} else {
  
  slgap
  
  }
```

There is a lack of affordable starter homes, with few units available at price points accessible to median-income workers in the region.

\[more analysis\]

### Homeownership Gaps

```{r}
## Homeownership Supply/Demand 

title_text <- "<span style = 'color:#8baeaa'><b>Supply</b></span> and <span style = 'color:#445ca9'><b>Demand</span> in the region for Homeowners"

ggplot(supply_demand_homeowner,
       aes(x = ami_order,
           y = estimate,
           fill = value)) +
  geom_col(position = "dodge") +
  theme_hda(base_size = 10) +
  scale_fill_hda() +
  coord_flip() +
  labs(title = title_text)

```

The existing supply skews towards larger, more expensive single-family homes, with sixty two percent of homeownership units cost more than \$1500 per month, and over 45,000 units have monthly costs exceeding \$2,000. In order to keep housing costs at 30% and below of income, a household would need to make at least \$80,000 per year to afford \$2000 in housing costs.

Only about 15,000 units in the region cost less than \$599 per month.

```{r}

#| label: fig-occ-price
#| fig.cap: "Maximum affordable home price by occupation versus median sales prices"

job_own <- ggplot(occ_locality_afford,
                   aes(y = reorder(group, price, decreasing = F),
                       x = price,
                       fill = type,
                       data_id = price,
                       tooltip = label_dollar()(price))) +
  geom_col() +
  geom_col_interactive(size = 2) +
  facet_grid(rows = vars(fct_rev(type)),
             scales = "free_y",
             switch = "y",
             space = "free_y") +
  scale_fill_hda() +
  scale_x_continuous(labels = label_dollar(), expand = expansion(mult = 0)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 24)) +
  labs(title = "Most workers cannot buy a home on their own",
       subtitle = "Maximum affordable home price by occupation<br>versus median sales prices",
       caption = "**Sources:** U.S. Bureau of Labor Statistics, Occupational Employment and Wage Statistics;<br>CVR MLS.") +
  theme_hda() +
  add_zero_line(axis = "x") +
  flip_gridlines() +
  theme(strip.text.y = element_blank())

if (knitr::is_html_output()) {
  
  girafe(ggobj = job_own,
         height_svg = 4) 
  
} else {
  
  job_own
  
  }

```

Median regional wages for the most commonly held jobs show a large gap in what purchase prices are affordable between occupations. Between the top five occupations, ranging from food preparation and service positions to management and business occupations, homes would need to be priced below \$100,000 ranging to over \$460,000 to be affordable.

Even for higher-paid public sector workers like firefighters, who could afford homes up to \$250,000 (assuming a 20% down payment), the options are limited as the market favors much higher-priced homes. The median home prices in the region far exceed what's affordable for many workers.

```{r}

local_wages_cty |> 
  ggplot(aes(x = aff_own, y = fct_reorder(locality, aff_own), label = label_dollar(accuracy = 1)(aff_own))) +
  geom_col(fill = hda_pal[1]) +
  geom_text(color = "white", hjust = 0.5, nudge_x = -15000) +
  #geom_vline(xintercept = 1620) +
  #geom_vline(xintercept = 1812) +
  #geom_text(
  #  aes(x = 1620, y = "King George", label = "Average\n1-bed\nrent:\n$1,620"), vjust = 0.6, hjust = 0, nudge_x = 20
  #) +
  #geom_text(
  #  aes(x = 1812, y = "King George", label = "Average\n2-bed\nrent:\n$1,812"), vjust = 0.6, hjust = 0, nudge_x = 20
  #) +
  scale_x_continuous(expand = c(0.01, 0.01), labels = label_dollar(accuracy = 1)) +
  scale_y_discrete(expand = c(0.1, 0)) +
  add_zero_line("x") +
  labs(
    title = "Affordable home purchase prices for public sector salaries",
    subtitle = "Based on average wages for police, fire, and teacher jobs",
    caption = "**Sources:** HDAdvisors calculations of local government salary data"
  ) +
  theme_hda(flip_gridlines = T)

```

For public sector workers to buy homes, they will likely need to move out of the region or pair up their incomes to increase their purchasing power. Using LEHD Origin-Destination Employment Statistics (LODES) data from the Census Bureau, commuter patterns can be seen for the region. This data allows for comparison between a respondent’s “origin” or residence, and their employment destination. In this analysis, commuters are individuals that travel outside of the county for work. Spotsylvania and Stafford have the highest number of commuters.

```{r}
# insert LODES viz
```
