# Regional spectrum {#sec-spectrum-reg}

This report outlines the *housing spectrum* concept and its utility for understanding housing market dynamics. It includes assessments of households in the region by various demographic and socioeconomic characteristics, primarily by wages/income. Special analysis is included for the region's core workforce, i.e. persons both living and working within the region.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

source("_common.R")

# Load packages other than tidyverse and hdatools
 
library(sf)
library(leaflet)
library(htmltools)

# Load data

hud_ami_areas <- read_rds("data/ami/hud_ami_areas.rds")

hud_ami_counties <- read_rds("data/ami/hud_ami_counties.rds") |> 
  mutate(
    hud_area_name = fct_reorder(hud_area_name, str_length(hud_area_name), .desc = TRUE)
  )

dc_msa <- read_rds("data/ami/dc_msa.rds")

faar_ami_all <- read_rds("data/ami/faar_ami_all.rds") |> 
  mutate(
    level = str_replace(level, "ami(\\d+)", "\\1% AMI"),
    area = str_replace_all(area, "HUD Metro FMR Area", "HMFR")
  )

fsa <- data.frame(
  np = 1:8,
  fsa = c("70%", "80%", "90%", "—", "108%", "116%", "124%", "132%")
  ) |> 
  pivot_wider(
    names_from = np,
    values_from = fsa
  )

mfi <- data.frame(
  area = unique(faar_ami_all$area),
  mfi = c(154700, 102800, 110400, 124000, 116300)
)

hmfrs <- unique(faar_ami_all$area)

ami_pal <- setNames(hda_pal[1:5], hmfrs[1:5])

```

## Area Median Income

With few exceptions, housing opportunities are primarily limited by how much money a household has available to buy or rent a home. To build a regional housing spectrum, we'll therefore start by grouping households according to their incomes. The most common approach uses a region's *Area Median Income* (AMI) as a reference point.

The U.S. Department of Housing and Urban Development (HUD) calculates AMIs each year using income estimates from the Census Bureau. Many housing programs managed by HUD, other public agencies, and nonprofit providers use AMIs to limit eligibility to households with low or moderate incomes. Because wages and living costs vary by market, AMIs help ensure housing assistance reaches those with the most needs.

### How AMIs are calculated

To calculate AMI, all households[^hh] are arranged from lowest to highest gross income. The household exactly in the middle has the *median* income for that given area. We can then refer to any other income value relative to that AMI. For example, if the AMI is \$80,000, a household with an income of \$40,000 is at 50% AMI.

[^hh]: As described further below, HUD's official income limits are actually derived from estimates of *family* incomes.

HUD's process to calculate AMIs for each area across the nation has two major parts:

1. Estimate the current *Median Family Income* (MFI)
2. Use MFI to generate Very Low-Income, Extremely Low-Income, and Low-Income limits 

::: {.callout-tip}

#### "Family" versus "Household"

HUD bases its AMIs on the median incomes of *families* rather than *households*. The Census Bureau definition for a family refers to a householder living with at least one other person related by birth, marriage, or adoption. Single persons living alone and households with *no* related individuals are excluded from MFI calculations. 

:::

#### Median Family Income

MFIs are calculated as follows:

1. **Data source:** HUD uses median family income data from the most recent version of the American Community Survey (ACS), an annual Census Bureau survey completed by 1% of the population. For FY 2024, the latest ACS estimates were from 2022.
2. **Reliability:** Because the ACS is based on a sample, HUD tests each the income estimates for each area to determine whether the margin of error is within an acceptable range.
3. **Inflation adjustment:** To account for the two-year lag from the ACS sample, HUD adjusts each value according to an inflation factor based on Consumer Price Index (CPI) projections published by the Congressional Budget Office (CBO).
4. **Round final value:** HUD rounds the adjusted median income value to the nearest $100. This result is the official MFI for a given area.

#### Income limit categories

To translate the MFI into income values that can be used to determine program eligibility, HUD follows a standard methodology to create new limits for three different income levels:

* **Extremely Low-Income:** Households whose incomes are not above 30% of the MFI
* **Very Low-Income:** Households whose incomes are not above 50% of the MFI
* **Low Income:** Households whose incomes are not above 80% of the MFI

For each of these, HUD generates specific values for household sizes between 1-person and 8-persons.

**Very Low-Income Limits**

1. Set 50% of the MFI as the preliminary 4-person Very Low-Income Limit (VLIL)
2. Adjustment VLIL up if:
    + Area has unusually high housing costs (35% of VLIL is no greater than 85% of annualized 2-bedroom FMR)
    + Area has unusually low incomes (VLIL is less than 50% of state's MFI for non-metro areas)
3. Adjust VLIL down if:
    + Area has unusually high incomes (VLIL is greater than 80% of U.S. MFI, and the annualized 2-bedroom FMR is no greater than 30% of VLIL)
4. Compare change with previous year and apply "ceiling or floor" rules:
    + Maximum annual increase: 10%
    + Minimum annual decrease: 5%
5. Adjust final VLIL by family size and round up to nearest $50

```{r}
#| label: tbl-fsa
#| tbl-cap: "Family size adjustments for number of persons in family"
#| html-table-processing: none

fsa |> 
  #mutate(across(1:8, ~ digits(.x, 2))) |> 
  kable(align = "cccccccc") |> 
  kable_styling(c("condensed"))

```

**Low-Income Limits**

1. Multiply the VLIL by 1.6 (80% divided by 50%) to get preliminary 4-person Low-Income Limit
2. Round to the nearest $50
3. Adjust down if preliminary value is greater than U.S. MFI ("capped")
4. Adjust up if 85% of annualized 2-bedroom FMR (multiplied by 1.6) is greater than 35% of preliminary value
5. Apply same "ceiling or floor" rules and family size adjustments used for Very Low-Income Limits

**Extremely Low-Income Limits**

1. Multiply the VLIL by 0.6 (30% divided by 50%) to get preliminary 4-person Extremely Low-Income Limit
2. Round to the nearest $50
3. Adjust by family size
4. Replace any values with federal poverty guidelines[^hhs] if guideline is higher

[^hhs]: As published by the Department of Health and Human Services (HHS): [U.S. Federal Poverty Guidelines](https://aspe.hhs.gov/topics/poverty-economic-mobility/poverty-guidelines) (aspe.hhs.gov)

### Income limit areas

HUD calculates AMIs for "income limit areas" that usually align with the Metropolitan Statistical Area (MSA) definitions published and occasionally revised by the U.S. Office of Management and Budget (OMB).[^omb] For FY 2024 AMIs, HUD based its geographies on the delineations provided in *OMB Bulletin No. 18-04* from September 2018.

[^omb]: Additional details on OMB definitions can be found here: [About Metropolitan and Micropolitan Areas](https://www.census.gov/programs-surveys/metro-micro.html) (census.gov)

However, HUD further subdivides many MSAs into "HUD Metro Fair Market Rent Areas" (HMFAs) that can sometimes be a single county.[^fmr] HFMAs can more accurately reflect income and market trends in a community, especially when they are part of very large metro areas like the Washington, D.C. region. If a county is not located in an MSA, HUD will calculate its unique AMI if reliable estimates are available.

[^fmr]: Fair Market Rents (FMRs) are also calculated by HUD at the same geographies it provides AMIs. FMRs are used to set payment standards for federal housing assistance programs. For more, see: [Fair Market Rents](https://www.huduser.gov/portal/datasets/fmr.html) (huduser.gov)

As of FY 2024, HUD uses four different income limit areas across the six localities included in this study. The table below shows which OMB and HUD designations correspond to each locality.

```{r}
#| label: tbl-ami-areas
#| tbl-cap: "OMB and HUD area designations by locality"
#| html-table-processing: none

hud_ami_areas |> 
  filter(GEOID %in% faar_fips) |> 
  mutate(
    msa = case_when(
      county %in% c("033", "099") ~ "Non-metro",
      .default = "Washington-Arlington-Alexandria, DC-VA-MD-WV Metro Area"
    )
  ) |> 
  arrange(desc(msa), desc(hud_area_name)) |> 
  select(msa, hud_area_name, County_Name) |> 
  kable(
    col.names = c("OMB Designation", "HUD Income Limit Area", "Locality")
  ) |> 
  column_spec(2, width = "250px") |>
  column_spec(3, width = "200px") |> 
  kableExtra::collapse_rows(columns = 1:2, valign = "top") |> 
  kableExtra::kable_styling(c("condensed"))

```

This map shows HUD income limit areas by color for localities in the Fredericksburg and Washington, D.C. regions. The Washington, D.C. MSA is outlined in black. 

```{css}

.quarto-figure-center>figure>p, .quarto-figure-center>figure>div {
    text-align: left;
}

```

```{r}
#| label: fig-ami-areas
#| fig-cap: "HUD FY 2024 Income Limit Areas in Fredericksburg and Washington, D.C. regions"
#| fig-asp: 0.8

hda_leaflet_colors <- colorFactor(
  palette = hda_pal_discrete(direction = 1)(4),  # Assuming you want to use all 6 colors
  domain = hud_ami_counties$hud_area_name
)

leaflet(options = leafletOptions(zoomControl = FALSE)) |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addMapPane("dc_msa", zIndex = 430) |> 
  addMapPane("faar_y", zIndex = 420) |> 
  addMapPane("faar_n", zIndex = 410) |> 
  addPolylines(
    data = dc_msa,
    color = "black",
    weight = 1.5,
    opacity = 1,
    options = pathOptions(pane = "dc_msa")
  ) |> 
  addPolygons(
    data = filter(hud_ami_counties, faar == "Y"),
    fillColor = ~ hda_leaflet_colors(hud_area_name),
    fillOpacity = 0.6,
    color = "white",
    opacity = 0.6,
    weight = 1.5,
    label = ~ as.character(County_Name),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "white",
      opacity = 1,
      fillOpacity = 0.7
    ),
    options = pathOptions(pane = "faar_y")
  ) |> 
  addPolygons(
    data = filter(hud_ami_counties, faar == "N"),
    fillColor = ~ hda_leaflet_colors(hud_area_name),
    fillOpacity = 0.3,
    color = "white",
    opacity = 0.4,
    weight = 1.5,
    label = ~ as.character(County_Name),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "white",
      opacity = 1,
      fillOpacity = 0.4
    ),
    options = pathOptions(pane = "faar_n")
  ) |> 
  addLegend(
    data = hud_ami_counties,
    pal = hda_leaflet_colors,
    values = ~hud_area_name,
    opacity = 0.8,
    title = "HUD FY 2024 Income Limit Areas",
    position = "topleft"
  ) |>
  addLegend(
    data = dc_msa,
    colors = c("black"),
    opacity = 0.9,
    labels = c("Washington-Arlington-Alexandria, DC-VA-MD-WV Metro Area"),
    position = "bottomleft"
  )

```

The *Washington-Arlington-Alexandria, DC-VA-MD-WV Metro Area* ("Washington, D.C. MSA") is a large MSA encompassing the capital and localities in three other states. HUD assigns much of this region to its *Washington-Arlington-Alexandria, DC-VA-MD HUD Metro FMR Area* --- including the City of Fredericksburg, Spotsylvania County, and Stafford County.

Although Culpeper County is part of the Washington, D.C. MSA, HUD considers it distinct enough to calculate its own AMI under the *Culpeper County, VA HUD Metro FMR Area*.

Because Caroline[^cc] and King George counties are not part of any MSA (i.e., non-metropolitan), HUD calculates their AMIs separately.

[^cc]: Caroline County was part of the Richmond, VA MSA until *OMB Bulletin No. 18-04*. HUD incorporated this change beginning with income limits published for FY 2022.

### AMIs in the Fredericksburg region

Median family incomes

```{r}
#| label: fig-mfi
#| fig-cap: "Median family incomes for HUD income limit areas"

mfi |> 
  filter(area != "Planning District 16") |> 
  mutate(
    area = fct_reorder(area, mfi),
    area = str_replace_all(area, "n-Al", "n-\nAl")  
  ) |> 
  ggplot(aes(x = mfi, y = area, label = label_dollar()(mfi))) +
  geom_col(fill = hda_pal[1]) +
  geom_text(
    color = "white",
    hjust = 1,
    nudge_x = -1500
  ) +
  scale_y_discrete(labels = label_wrap(24)) +
  scale_x_continuous(
    expand = c(0.01, 0.01),
    labels = label_dollar()
    ) +
  add_zero_line("x") +
  labs(
    title = "Median family incomes for HUD income limit areas",
    subtitle = "FY 2024 values based on 2022 American Community Survey",
    caption = "**Source:** HUD Office of Policy Development and Research, FY 2024 Income Limits."
  ) +
  theme_hda(flip_gridlines = TRUE)

```

AMIs by family size

```{r}
#| label: fig-ami
#| fig-cap: "Area Median Incomes by family size for HUD income limit areas"

hud_ami_plot <- faar_ami_all |> 
  filter(
    area != "Planning District 16",
    #str_detect(hh_size, "[1234]" ),
    str_detect(level, "[358]")
  ) |> 
  filter(
    !(str_starts(area, "W|K") & cap == "Uncapped" & level == "80% AMI")
  ) |> 
  group_by(area, hh_size) |> 
  mutate(area = str_replace_all(area, "n-Al", "n-\nAl")) |> 
  mutate(
    income_plot = case_when(
      level == "30% AMI" ~ income,
      .default = income - lag(income)
    )
  )

label <- data.frame(
  hh_size = 1,
  income = c(97800, 113700),
  level = "80% AMI",
  area = "Caroline County, VA",
  label = c("U.S. MFI: **$97,800**", "Virginia<span> </span>MFI: **$113,700**")
)

fig_ami_subtitle <- paste(
  "FY 2024 values for",
  "<span style='color: #e9ab3f'><b>30% AMI</b></span>,",
  "<span style='color: #8baeaa'><b>50% AMI</b></span>, and",
  "<span style='color: #445ca9'><b>80% AMI</b></span>"
)

hud_ami_plot |> 
  ggplot(aes(x = str_sub(hh_size, 1, 1), y = income_plot, fill = rev(level))) +
  geom_hline(yintercept = 97800) +
  geom_hline(yintercept = 113700, linetype = 2) +
  geom_col(position = "stack", alpha = 0.6, width = 1) +
  geom_segment(
    aes(
      x = as.numeric(factor(hud_ami_plot$hh_size)) - 0.5,
      xend = as.numeric(factor(hud_ami_plot$hh_size)) + 0.5,
      y = income,
      yend = income,
      color = level
    ),
    linewidth = 2
  ) +
  geom_richtext(
    data = label,
    aes(x = hh_size, y = income, label = label),
    family = "Lato",
    size = 3,
    hjust = 0,
    nudge_x = -0.6,
    nudge_y = 5000,
    fill = NA,
    label.color = NA
  ) +
  facet_wrap(~area, nrow = 1, labeller = labeller(area = markdown_wrap_gen(16))) +
  scale_fill_hda(guide = "none") +
  scale_color_hda(-1) +
  scale_y_continuous(
    limits = c(0, 140000),
    breaks = seq(0, 120000, 20000),
    expand = c(0.01, 0.05),
    labels = label_dollar()
  ) +
  labs(
    title = "Area Median Incomes by family size for HUD income limit areas",
    subtitle = fig_ami_subtitle,
    caption = "**Source:** HUD Office of Policy Development and Research, FY 2024 Income Limits."
  ) +
  add_zero_line("y") +
  theme_hda()

```


```{r}

hud_ami_uncap <- faar_ami_all |> 
  filter(
    area != "Planning District 16",
    str_detect(level, "[358]")
  ) |> 
  filter(
    !(str_starts(area, "W|K") & cap == "Uncapped" & level == "80% AMI")
  ) |> 
  group_by(area, hh_size) |> 
  mutate(
    income_plot = case_when(
      level == "30% AMI" ~ income,
      .default = income - lag(income)
    )
  )

```

